---
title: "AFM-MCD Robust T-squared Control Chart: A Quick Start for Quality Engineers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AFM-MCD Robust T-squared Control Chart: A Quick Start for Quality Engineers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi       = 96,
  warning   = FALSE,
  message   = FALSE
)
```

This vignette is deliberately short. It shows the four-line pipeline
a quality engineer runs to deploy the AFM-MCD Phase 2 control chart,
why the robust version matters when historical batches are not perfectly
clean, and what to do when the estimator complains. The method behind
the code is Frutos-Galarza et al. (2026); see the package references
for the full derivation.

## 1. Why not just use classical Hotelling?

Classical Hotelling T-squared assumes the Phase 1 calibration batches are
in control. In practice, one or two historical batches almost always
carry within-batch outliers or a shifted mean. Those pollute the pooled
covariance matrix S_p, which then inflates the UCL. The chart looks
"quiet" but the increased UCL is masking real Phase 2 alarms.

The AFM-MCD chart absorbs this contamination at calibration time:

- **MCD per batch** trims the ~33% most extreme observations in each
  batch (`h = 0.67`), so within-batch outliers do not enter the batch
  covariance S_k.
- **AFM inverse weights** `w_k = (1/lambda1_k) / Σ (1/lambda1_i)`
  down-weight batches with unusually large dispersion, so entirely
  shifted batches contribute less to the reference covariance S_w.

Phase 2 then uses simple batch means against the frozen reference
`(mu_r, S_w)`, without any per-batch trimming. The alarm sensitivity
of the classical chart is preserved, but the masking is gone.

## 2. Same data, two charts

To make the point concrete, we simulate a Phase 1 that already contains
2 batches with within-batch outliers and ~7 % of batches fully shifted,
plus a Phase 2 where 30 % of batches are out of control. We then run
the classical chart and the AFM-MCD chart on the **same** data.

```{r sim}
library(robustT2AFM)

sim <- simulate_batch_process(
  K1 = 30, K2 = 20, I = 20, J = 4,
  outlier_batches_F1 = 2, outlier_rate = 0.20, outlier_shift = 4,
  prop_contam_F1 = 0.07, shift_contam = 3,
  prop_ooc_F2 = 0.30,   shift_ooc = 2,
  seed = 20260417
)
vars   <- paste0("Var", 1:4)
phase1 <- subset(sim, Phase == "Phase 1")
phase2 <- subset(sim, Phase == "Phase 2")
```

Classical Hotelling (baseline):

```{r classical}
cal_c <- hotelling_classical_calibrate(phase1, vars)
mon_c <- hotelling_classical_monitor(phase2, cal_c, vars)
ucl_c <- hotelling_classical_ucl(K = 30, I = 20, J = 4)
plot_control_chart(
  mon_c, UCL = ucl_c$UCL, alpha = ucl_c$parameters$alpha,
  method_label = "Classical Hotelling T-squared - Phase 2"
)
```

AFM-MCD (this package):

```{r afm-mcd}
cal_r <- calibrate_afm_mcd(phase1, vars)
ucl_r <- ucl_F_adjusted(cal_r, I = 20)
mon_r <- monitor_afm_mcd(phase2, cal_r, vars)
plot_control_chart(
  mon_r, UCL = ucl_r$UCL, alpha = ucl_r$parameters$alpha,
  method_label = "AFM-MCD Robust T-squared - Phase 2"
)
```

The two Phase 2 datasets are identical. The AFM-MCD chart flags the
out-of-control batches; the classical chart, with its UCL inflated by
the contaminated Phase 1, misses several of them.

## 3. The pipeline on your own data

Your Phase 1 and Phase 2 data need to sit in a single `data.frame` with:

- a `Batch` column identifying each batch (character or factor),
- one numeric column per process variable (`Var1`, `Var2`, ...).

In production you load them with `read.csv()` or your MES connector.
In this vignette we keep the simulator so the code is reproducible:

```r
# In practice, replace this line with your data load:
#   raw <- read.csv("historical_batches.csv")
# and split into phase1 / phase2 as you already do.
raw <- simulate_batch_process(seed = 20260417)
```

Then the entire chart is four operational calls:

```r
vars <- paste0("Var", 1:4)
cal  <- calibrate_afm_mcd(subset(raw, Phase == "Phase 1"), vars)
ucl  <- ucl_F_adjusted(cal, I = 20)
mon  <- monitor_afm_mcd(subset(raw, Phase == "Phase 2"), cal, vars)
plot_control_chart(mon, UCL = ucl$UCL, alpha = ucl$parameters$alpha,
                   method_label = "AFM-MCD Control Chart - Phase 2")
```

## 4. Reading the chart and putting it in a report

- **Top-left summary** ("N OOC / total (X%)"): how many Phase 2 batches
  crossed the UCL and what fraction of the run that is.
- **UCL label** ("UCL = 19.7  (alpha = 0.001)"): the limit itself and
  the nominal false-alarm rate, so anyone reading the report knows the
  chart's ARL0.
- **Red batch labels** (e.g. `F2_B07`): the batches to investigate.
  Pull their raw records, check the process log around that time
  (materials, operator, equipment), and if the shift persists across
  successive batches, escalate.
- **Table of alarms:**

```r
mon$is_ooc <- mon$T2 > ucl$UCL
mon[mon$is_ooc, ]
```

To include the chart in a report, add `save_path` and you get both a
300 dpi PNG and a vector PDF at 7 x 4.5 in:

```r
plot_control_chart(mon, UCL = ucl$UCL, alpha = ucl$parameters$alpha,
                   method_label = "AFM-MCD Control Chart - Phase 2",
                   save_path = "phase2_chart")
# writes phase2_chart.png and phase2_chart.pdf
```

## 5. Troubleshooting

**Too few observations per batch for MCD.** MCD needs enough rows per
batch to fit a trimmed covariance. Practical minimum: `I >= J + 2`
(e.g. `J = 4` implies `I >= 6`); the package aborts calibration if
fewer than two batches survive. Effective batch size is `m* = round(I*h)`
with `h = 0.67`, so the F-adjusted UCL loses degrees of freedom when I
is small. Rule of thumb: aim for `m* >= 2 * J`. If your history is
thin (say `K1 < 5` or `I < 8`), the UCL becomes conservative and
unstable; collect more calibration data before going live.

**Variables on very different scales.** MCD itself is
affine-equivariant, so per-batch estimation is unaffected by scale
differences. But the AFM weights rank batches by their largest
eigenvalue, so a variable with much larger natural variance will
dominate `lambda1` and drive the weights. Standardize using **Phase 1
statistics only**, then apply the same shift and scale to Phase 2:

```r
mu_p1 <- colMeans(phase1[vars])
sd_p1 <- apply(phase1[vars], 2, sd)
phase1[vars] <- scale(phase1[vars], center = mu_p1, scale = sd_p1)
phase2[vars] <- scale(phase2[vars], center = mu_p1, scale = sd_p1)
```

Never restandardize a new batch with its own mean and SD - that erases
the mean shift you are trying to detect.

**S_w looks singular in Phase 2.** When two variables are almost
collinear (correlation above ~0.99), `S_w` can be numerically singular
and `monitor_afm_mcd()` fails when it tries to invert it. Two fixes:
drop the redundant variable, or increase I so that
`m* = round(I * 0.67) > J` with margin (again, `m* >= 2 * J` is a good
target).

## Reference

Frutos-Galarza, S. D., Ruiz-Barzola, O., Ramírez, J., Galindo-Villardón, P.
(2026). *A Robust Hotelling-Type T2 Control Chart Combining the Minimum
Covariance Determinant Estimator with Multiple Factor Analysis Weighting.*
Manuscript in preparation.
