---
title: "AFM-MCD Robust T-squared Control Chart: A Quick Start for Quality Engineers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AFM-MCD Robust T-squared Control Chart: A Quick Start for Quality Engineers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi       = 96,
  warning   = FALSE,
  message   = FALSE
)
```

This vignette is deliberately short. It shows the one call a quality
engineer runs to deploy the AFM-MCD Phase 2 control chart, why the
robust version matters when historical batches are not perfectly clean,
the four steps behind that call for anyone who wants them, and what to
do when the estimator complains. The method behind the code is
Frutos-Galarza et al. (2026); see the package references for the full
derivation.

## 1. Start here

The package ships the example data of the paper's base configuration, so
this runs as it stands:

```{r start}
library(robustT2AFM)

study <- run_afm_mcd(afm_phase1, afm_phase2)
study
```

`afm_phase1` holds 30 historical batches, 6 of them contaminated;
`afm_phase2` holds 20 batches to monitor, half of them off-target.
Replace them with your own data when you have it: each needs a column
identifying the batch and one numeric column per process variable. If
that column is not called `Batch`, say so with `batch_col = "..."`
rather than renaming your data.

`summary(study)` prints the full report, and `study$chart` is the
control chart:

```{r start-chart, fig.alt = "AFM-MCD control chart for Phase 2"}
study$chart
```

## 2. Why not just use classical Hotelling?

Classical Hotelling T-squared assumes the Phase 1 calibration batches are
in control. In practice, one or two historical batches almost always
carry within-batch outliers or a shifted mean. Those pollute the pooled
covariance matrix `Sp`. The control limit itself does not move:
`hotelling_classical_ucl()` is parametric and never sees `Sp`. What
moves is the statistic. Since T-squared divides by `Sp`, an inflated
`Sp` shrinks every distance at once, and batches that would have
crossed a fixed limit no longer reach it. On this data the classical
chart flags 9 of 20 when Phase 1 is clean and 2 when it is
contaminated, against the same limit of 19.4644.

The AFM-MCD chart absorbs this contamination at calibration time:

- **MCD per batch** trims the ~33% most extreme observations in each
  batch (`h = 0.67`), so within-batch outliers do not enter the batch
  covariance S_k.
- **AFM inverse weights** `w_k = (1/lambda1_k) / Σ (1/lambda1_i)`
  down-weight batches with unusually large dispersion, so entirely
  shifted batches contribute less to the reference covariance S_w.

Phase 2 then uses simple batch means against the frozen reference
`(mu_r, S_w)`, without any per-batch trimming. The alarm sensitivity
of the classical chart is preserved, but the masking is gone.

## 3. Same data, two charts

To make the point concrete, we run the classical chart and the AFM-MCD
chart on the **same** batches. `afm_phase1` has 6 of its 30 batches
carrying 4 outlying observations each, shifted 4 sigma; half the
batches of `afm_phase2` are off-target by 1 sigma.

```{r sim}
vars   <- paste0("Var", 1:4)
phase1 <- afm_phase1
phase2 <- afm_phase2
```

Both objects also carry `Status` and `ContaminationType`, which say
which batches are contaminated. They exist because these are simulated
data. Your own historical data will not have them, and the method does
not need them: which batches are out of control is the question the
chart answers, not an input it requires. Use
`simulate_batch_process()` if you want a different scenario.

Both charts on the same batches, in the same order, each against its
own limit:

```{r comparison, fig.height=4.5}
cal_c <- hotelling_classical_calibrate(phase1, vars)
ucl_c <- hotelling_classical_ucl(K = 30, I = 20, J = 4)
mon_c <- hotelling_classical_monitor(phase2, cal_c, vars, ucl = ucl_c$UCL)

cal_r <- calibrate_afm_mcd(phase1, vars)
ucl_r <- ucl_F_adjusted(cal_r, I = 20)
mon_r <- monitor_afm_mcd(phase2, cal_r, vars, ucl = ucl_r$UCL)

plot_method_comparison(
  robust = mon_r, UCL_robust = ucl_r$UCL,
  classical = mon_c, UCL_classical = ucl_c$UCL
)
```

The two Phase 2 datasets are identical. With this seed the AFM-MCD
chart flags 8 of the 10 off-target batches and the classical chart
flags 2, and neither raises a false alarm on the 10 in-control
batches.

The ninth is F2_B14, a genuinely faulty batch at 0.97 of the limit.
The count is 8 of 10, and the one that got away was three per cent
short of firing.

Three cautions about those counts, so they are not over-read:

- **They are one replication, not detection rates.** The paper reports
  0.863 for the proposed method and 0.270 for the classical one,
  averaged over 2000 replications. Re-running the block with a
  different seed will give different counts.
- **The paper measures them differently.** There both methods are
  first equalised to a common in-control ARL0 using the empirical
  quantile of each method's own null distribution over 5000
  in-control batches; here each method uses its own operational
  limit. The two figures describe similar but not identical
  quantities, and neither estimates the other.
- **This shows half the argument.** With a 1-sigma shift and
  `alpha = 0.001`, what separates the two methods here is detection
  power. Across twelve seeds and 200 replications neither method
  produced a single false alarm, so nothing in this comparison speaks
  to false-alarm behaviour.

## 4. Your own data, and the four steps behind the one call

Your Phase 1 and Phase 2 data need to sit in `data.frame`s with:

- a column identifying each batch (character or factor), called `Batch`
  unless you say otherwise with `batch_col`,
- one numeric column per process variable.

In production you load them with `read.csv()` or your MES connector:

```r
raw    <- read.csv("historical_batches.csv")
phase1 <- subset(raw, Phase == "Phase 1")
phase2 <- subset(raw, Phase == "Phase 2")

study  <- run_afm_mcd(phase1, phase2, batch_col = "Lote")
```

`run_afm_mcd()` recomputes nothing of its own: it calls the four
functions below in order and keeps their results untouched. Call them
directly when you want the pieces, which is how the paper's own
scripts are written:

```r
vars <- paste0("Var", 1:4)
cal  <- calibrate_afm_mcd(afm_phase1, vars)          # Phase 1 reference
ucl  <- ucl_F_adjusted(cal, I = 20)                  # control limit
mon  <- monitor_afm_mcd(afm_phase2, cal, vars,       # Phase 2 statistics
                        ucl = ucl$UCL)
plot_control_chart(mon, UCL = ucl$UCL, alpha = ucl$parameters$alpha,
                   method_label = "AFM-MCD Control Chart - Phase 2")
```

## 5. Reading the chart and putting it in a report

- **Top-left summary** ("N OOC / total (X%)"): how many Phase 2 batches
  crossed the UCL and what fraction of the run that is.
- **UCL label** ("UCL = 19.7  (alpha = 0.001)"): the limit itself and
  the nominal false-alarm rate, so anyone reading the report knows the
  chart's ARL0.
- **Red batch labels** (e.g. `F2_B12`): the batches to investigate.
  Pull their raw records, check the process log around that time
  (materials, operator, equipment), and if the shift persists across
  successive batches, escalate.
- **Table of alarms:** passing `ucl` to `monitor_afm_mcd()` above added an
  `is_ooc` column, flagging every batch strictly above the limit (a batch
  landing exactly on it counts as under control, the same criterion the
  chart uses):

```r
mon[mon$is_ooc, ]
```

To include the chart in a report, add `save_path` and you get both a
300 dpi PNG and a vector PDF at 7 x 4.5 in:

```r
plot_control_chart(mon, UCL = ucl$UCL, alpha = ucl$parameters$alpha,
                   method_label = "AFM-MCD Control Chart - Phase 2",
                   save_path = "phase2_chart")
# writes phase2_chart.png and phase2_chart.pdf
```

## 6. Troubleshooting

**Too few observations per batch for MCD.** MCD needs enough rows per
batch to fit a trimmed covariance. Practical minimum: `I >= J + 2`
(e.g. `J = 4` implies `I >= 6`); the package aborts calibration if
fewer than two batches survive. Effective batch size is `m* = round(I*h)`
with `h = 0.67`, so the F-adjusted UCL loses degrees of freedom when I
is small. Rule of thumb: aim for `m* >= 2 * J`. If your history is
thin (say `K1 < 5` or `I < 8`), the UCL becomes conservative and
unstable; collect more calibration data before going live.

**Variables on very different scales.** MCD itself is
affine-equivariant, so per-batch estimation is unaffected by scale
differences. But the AFM weights rank batches by their largest
eigenvalue, so a variable with much larger natural variance will
dominate `lambda1` and drive the weights. Standardize using **Phase 1
statistics only**, then apply the same shift and scale to Phase 2:

```r
mu_p1 <- colMeans(phase1[vars])
sd_p1 <- apply(phase1[vars], 2, sd)
phase1[vars] <- scale(phase1[vars], center = mu_p1, scale = sd_p1)
phase2[vars] <- scale(phase2[vars], center = mu_p1, scale = sd_p1)
```

Never restandardize a new batch with its own mean and SD - that erases
the mean shift you are trying to detect.

**S_w looks singular in Phase 2.** When two variables are almost
collinear (correlation above ~0.99), `S_w` can be numerically singular
and `monitor_afm_mcd()` cannot invert it. It stops with an error that
names the likely cause and what to do, rather than the raw message from
`solve()`. The two fixes it suggests: drop the redundant variable, or
increase I so that `m* = round(I * 0.67) > J` with margin (again,
`m* >= 2 * J` is a good target).

A constant column produces the same failure, and that is worth knowing
because it is how an identifier column usually announces itself. A
column that varies but is not a measurement — a sample index, a run
number — will *not* fail: it goes through as a process variable and the
study completes, looking entirely credible. That is why
`run_afm_mcd()` always reports which variables it detected when you do
not name them.

**Batches of different sizes.** Plant data rarely has equal batches:
one run stops early, another loses discarded rows. Calibration copes
with that, but the control limit assumes a single batch size I, so
there is no exact limit for such a calibration.
`calibrate_afm_mcd()` warns and records `I_phase1` as the size whose
`m*` equals the average of the per-batch `m*`, which is what matches
the degrees of freedom of the covariance estimate. The full vector is
in `calibration$batch_sizes`. To choose the value yourself, pass it:

```r
ucl <- ucl_F_adjusted(cal, I = 18)
```

## Reference

Frutos-Galarza, S. D., Ruiz-Barzola, O., Ramírez, J., Galindo-Villardón, P.
(2026). *A Robust Hotelling-Type T2 Control Chart Combining the Minimum
Covariance Determinant Estimator with Multiple Factor Analysis Weighting.*
Manuscript in preparation.
